\myTitle

### 2023-11-05 body-text-of-paper.tex
matt's updates to methods and results

In the \gscramble{} package there is a convenience function called {\footnotesize\tt create\_GSP()} that
will create GSPs for samples up to and including second-generation backcrosses (an example appears in Figure~\ref{fig:exampleGSPs}); 
%%%%%
\begin{figure}
\newcommand{\gspcaptwo}{\footnotesize
An example GSP generated by {\footnotesize\tt create\_GSP()} that produces four different
admixture categories: one~$F_1$ in sample node~s7; two~$F_2$'s in sample node~s11; one $BC_1$ in sample node~s9; and two $BC_2$'s in sample node~s10. 
}
\begin{center}
\includegraphics[width=\columnwidth]{figures/AllThem.pdf}
\end{center}
\caption[\gspcaptwo]{\gspcaptwo}
\label{fig:gsp1}
\end{figure}
%%%%
however, \gscramble{} is designed
to handle user-specified GSPs of arbitrary complexity.  When creating one's own GSP, it is useful
to keep in mind a handful of requirements for a GSP to be valid. Writing $\founders$ for the set of all founder nodes, $\nonfound$ for the set of all non-founder nodes, and $\nonfounds$ for all non-founder
nodes that are not parents of other non-founders, but are adjacent to a sample node, $\nonfounde$ for all
non-founder nodes that have non-founder node daughters, but no sample nodes, and $\nonfoundse$ for all
non-founder nodes that are parents of non-founder nodes and also adjacent to sample-nodes, the necessary conditions
for a valid GSP, in addition to having no inbreeding loops, are:
\begin{equation}
\begin{aligned}
\sum_{x \in \nonfounds \cup \nonfoundse} S_x &= |\founders| &  \\
\sum_{i=1}^{e_x} G^-_{x,i} &= 2 & \forall x \in \founders \\
G^+_{x,1} &= G^+_{x,2} & \forall x \in \nonfound \\
\sum_{i=1}^{e_x} G^-_{x,i} &= G^+_{x,1} +  G^+_{x,2} & \forall x \in \nonfounde  \\
 S_x/2 + \sum_{i=1}^{e_x} G^-_{x,i} &= G^+_{x,1} +  G^+_{x,2} & ~~~\forall x \in \nonfoundse.  \\
\end{aligned}
\end{equation}
Before simulating from a GSP, \gscramble{} checks to ensure the GSP
satisfies these validity conditions.  The simulation of genomic segments
from founders to samples in a GSP is performed by the \gscramble{} function
{\footnotesize\tt segregate()}, which requires as input a list of GSPs through which
to segregate genomic segments, and, for each such GSP, a table that maps the
population labels on each GSP (in the inverted triangles atop the founders) to the
different sampled groups of individuals.

@@

These permutations are executed upon the genotypes provided by the user.
\gscramble{} accepts genetic data as a matrix with $N$ rows (one for each individual) and
$2L$ columns (two for each locus). If the data have been phased, then the phased alleles from the
first haplotype should occupy the
first entry for each locus within an individual, while alleles phased together on the second hapltotype should occupy the
second entry for each locus. 
There can be several populations represented in the data set, but
each individual must belong to exactly one
population.

@@

\subsubsection*{Feral pigs in Missouri}

Genetic samples were collected from invasive wild pigs across a SPATIAL EXTENT in southeastern Missouri that were lethally removed through population control efforts conducted as a component of the Missouri Feral Hog Elimination Partnership
by US Department of Agriculture – Animal Plant Health Inspection Services – Wildlife Services, Missouri Department of Conservation, and other cooperative state agencies CHECK.
DNA was extracted from hair or tissue (collected from animals at the time of euthanasia) with commercially available magnetic bead recovery kits (MagMax DNA, Thermo Fisher Scientific).
Genotyping was performed with GeneSeek’s Genomic Profiler for Porcine HD which provides 62,128 biallelic SNP loci that are mapped across the 18 autosomal chromosomes.
We implemented standard genotype quality control metrics, specifically, we pruned genotypes with call rates <95\% and minor allele frequencies <5\%.
We implemented linkage disequilibrium (LD) pruning using a window size of 50 loci and step size of 5 loci to remove markers above a linkage threshold of R2 > 0.5.

@@

@@

\section*{Acknowledgements}
We thank etc. etc.   Contribution number  mHAVEAGAS-003

>>>>

In the \gscramble{} package there is a convenience function called {\footnotesize\tt create\_GSP()} that
will create GSPs for samples up to and including second-generation backcrosses (an example appears in Figure~\ref{fig:gsp2}); 
%%%%%
\begin{figure}
\newcommand{\gspcaptwo}{\footnotesize
An example GSP generated by {\footnotesize\tt create\_GSP()} that produces four different
admixture categories: one~$F_1$ in sample node~s7; two~$F_2$'s in sample node~s11; one $BC_1$ in sample node~s9; and two $BC_2$'s in sample node~s10. 
}
\begin{center}
\includegraphics[width=\columnwidth]{figures/AllThem.pdf}
\end{center}
\caption[\gspcaptwo]{\gspcaptwo}
\label{fig:gsp2}
\end{figure}
%%%%
however, \gscramble{} is designed
to handle user-specified GSPs of arbitrary complexity.  When creating one's own GSP, it is useful
to keep in mind a handful of requirements for a GSP to be valid. Writing $\founders$ for the set of all founder nodes, $\nonfound$ for the set of all non-founder nodes, and $\nonfounds$ for all non-founder
nodes that are not parents of other non-founders, but are adjacent to a sample node, $\nonfounde$ for all
non-founder nodes that have non-founder node daughters, but no sample nodes, and $\nonfoundse$ for all
non-founder nodes that are parents of non-founder nodes and also adjacent to sample-nodes, the necessary conditions
for a valid GSP, in addition to having no inbreeding loops, are:
\begin{equation}
\begin{aligned}
\sum_{x \in \nonfounds \cup \nonfoundse} S_x &= |\founders| &  \\
\sum_{i=1}^{e_x} G^-_{x,i} &= 2 & \forall x \in \founders \\
G^+_{x,1} &= G^+_{x,2} & \forall x \in \nonfound \\
\sum_{i=1}^{e_x} G^-_{x,i} &= G^+_{x,1} +  G^+_{x,2} & \forall x \in \nonfounde  \\
 S_x/2 + \sum_{i=1}^{e_x} G^-_{x,i} &= G^+_{x,1} +  G^+_{x,2} & ~~~\forall x \in \nonfoundse.  \\
\end{aligned}
\end{equation}
Before simulating from a GSP, \gscramble{} checks to ensure the GSP
satisfies these validity conditions.  The simulation of genomic segments
from founders to samples in a GSP is performed by the \gscramble{} function
{\footnotesize\tt segregate()}, which requires as input a list of GSPs through which
to segregate genomic segments, and, for each such GSP, a table that maps the
population labels on each GSP (in the inverted triangles atop the founders) to the
different sampled groups of individuals.

@@

These permutations are executed upon the genotypes provided by the user.
\gscramble{} accepts genetic data as a matrix with $N$ rows (one for each individual) and
$2L$ columns (two for each locus). If the data have been phased, then the phased alleles from the
first haplotype should occupy the
first entry for each locus within an individual, while alleles phased together on the second haplotype should occupy the
second entry for each locus. 
There can be several populations represented in the data set, but
each individual must belong to exactly one
population.

@@

\subsubsection*{Invasive wild pigs in Missouri}

Genetic samples were collected from 160 invasive wild pigs across a SPATIAL EXTENT in 
southeastern Missouri that were lethally removed through population control efforts conducted 
as a component of the Missouri Feral Hog Elimination Partnership
by US Department of Agriculture – Animal Plant Health Inspection Services – Wildlife Services, Missouri Department of Conservation, 
and other cooperative state agencies CHECK.
DNA was extracted from hair or tissue (collected from animals at the time of euthanasia) with 
commercially available magnetic bead recovery kits (MagMax DNA, Thermo Fisher Scientific).
Genotyping was performed with GeneSeek’s Genomic Profiler for Porcine HD which provides 
62,128 biallelic SNP loci that are mapped across the 18 autosomal chromosomes.
We implemented standard genotype quality control metrics, specifically, we pruned genotypes with 
call rates <95\% and minor allele frequencies <5\%.
We implemented linkage disequilibrium (LD) pruning using a window size of 50 loci and step size 
of 5 loci to remove markers above a linkage threshold of $R^2 > 0.5$.

The 160 invasive wild pig samples represented 120 reference individuals from the core regions of 
3 populations (40 individuals each from Pop1, Pop2, Pop14) and 40 individuals from the contact regions of these populations. 
We used ADMIXTURE version 1.3.0 \citep{alexander2009fast} to characterize the genetic structure of the 160 samples (for $K = {1...6}$)
and identify that indeed the 120 reference individuals represented 3 distinct genetic clusters.

Using the 120 reference individuals, we sought to classify the admixture patterns of the 40 empirical samples 
from the contact regions using simulated pedigrees from \gscramble{}. To address this objective we simulated 
pedigrees to represent the likely classes of admixed individuals that would be present in the samples 
from the contact regions of the 3 reference populations. The types of admixed individuals we considered were 
all two-population combinations of $F_1$, $F_2$, $B_1$ and $B_2$. We also 
simulated individuals from each of the founder populations. The first step of this workflow involved 
simulating admixed individuals for 2 founder populations using one of the preset \gscramble{} 
GSPs provided by the function {\footnotesize\tt create\_GSP()} for producing each of the types of admixed 
individuals. While we could have simulated all 4 types of admixed individuals from a single GSP (Figure~\ref{fig:gsp2}), 
different numbers of simulated individuals are produced for each hybrid class. Therefore, for the sake of the 
computational efficiency from simulating the exact number of desired individuals, we chose to simulate from 
a single GSP for each of the hybrid classes by setting the corresponding hybrid parameter in {\footnotesize\tt create\_GSP()} to {\footnotesize\tt TRUE}. We used the default settings of the {\footnotesize\tt plink\_map2rec\_rates()} to specify the recombination rates. The 120 simulated genotypes were written to PLINK .ped and .map files using {\footnotesize\tt gscramble2plink()}. We used PLINK version 1.9 \citep{purcell2007plink} to convert the genotypes to binary files and used these as input into ADMIXTURE for which we specified $K=3$ to produce ancestry values. We iteratively 
performed this workflow to produce 1,000 simulated genotypes and ancestry values for each of the 
4 hybrid classes, as well as founder individuals, for each of the population combinations, for a total of 
21,000 simulated individuals.

@@

\subsubsection*{Invasive wild pigs in Missouri}

SNP filtering resulted in 14,233 loci that were used for subsequent analysis. Using the 160 empirical samples, ADMIXTURE results supported the identification of 3 genetic clusters (SUPPLEMENTAL FIGURE CROSS VALIDATION?). Admixture *q* values clearly grouped individuals from each of the core populations into 3 clusters while the 40 individuals from the contact regions were highly admixed (SUPPLEMENTAL FIGURE ADMIXTURE). The
$F_{ST}$ divergence from ADMIXTURE was: $F_{ST} = 0.243$ for Pop1-Pop2, $F_{ST} = 0.145$ for Pop1-Pop14 and $F_{ST} = 0.133$ for Pop2-Pop14.

@@

\section*{Acknowledgements}
We thank etc. etc.   Contribution number  mHAVEAGAS-003. 
This work utilized the Alpine high performance computing resource at the University of Colorado Boulder. Alpine is jointly funded by the University of Colorado Boulder, the University of Colorado Anschutz, Colorado State University, and the National Science Foundation (award 2201538).
